---
title: "water-balance-simulation"
output:
  html_document:
  pdf_document: default
date: "`r Sys.Date()`"
---
## Summary
This model simulates monthly water balance in a rainwater harvesting system. The example uses rainfall data for Tororo District, Uganda as an example for the method. At the end, application of the approach to other datasets and other analysis of the data is considered.
```{r echo=FALSE}
library(tidyverse)
```
## Data
The data is taken from World Food Prorgamme (WFP) satellite data, publicly available through [Humanitarian Data Exchange](https://data.humdata.org/dataset/uga-rainfall-subnational).

Variables provided in the source data are as follows:

| Source  | Description                                            | Unit |
|---------|--------------------------------------------------------|------|
| rfh     | 10 day rainfall                                        | mm   |
| r1h     | Rainfall 1-month rolling aggregation                   | mm   |
| r3h     | Rainfall 3-month rolling aggregation                   | mm   |
| rfh_avg | Rainfall long term average                             | mm   |
| r1h_avg | Rainfall 1-month rolling aggregation long term average | mm   |
| r3h_avg | Rainfall 3-month rolling aggregation long term average | mm   |
| rfq     | Rainfall anomaly                                       | \%   |
| r1q     | Rainfall 1-month anomaly                               | \%   |
| r3q     | Rainfall 3-month anomaly                               | \%   |

## Preparation
The data can be loaded from either the online source or the downloaded file. The code below reads the file and glimpses the data to check its contents. 
```{r echo=TRUE}
# Use the url to read the file available online, or: 
  # Download the rainfall .csv from link above,
  # Delete "url," from the call to read_csv(...) below,
  # Uncomment the call to file.choose(),
  # Select the .csv containing rainfall data from downloads on your machine when prompted.

url <- 'https://data.humdata.org/dataset/5ce9dfd7-b391-48cd-ad22-a6d0769fafe7/resource/96d0169f-0b56-428b-b59e-27e82c3f4eff/download/uga-rainfall-adm2-full.csv'

data <- #file.choose() |>
  read_csv(url, col_types = cols("D", "i", "c", "d", "d", "d", "d", "d", "d", "d", "d", "d", "d", "c")) |>
  # Remove the top row with no useful information. 
  slice(-1)

glimpse(data)
```
The overview of the data shows a few problems to be addressed:

* The data is not tidy:
  - There are NA values in the aggregate measures of rainfall,
  - Measurements over different time periods are provided in one dataset. 
* The location data (P-Code) is not human-readable.

Named P-Codes are available on HDX, so we can read these in and add them to the data.
``` {r echo=TRUE}
# Set the URL of the file containing named P-Codes. 
url_pcode <- 'https://data.humdata.org/dataset/cb963915-d7d1-4ffa-90dc-31277e24406f/resource/f65bc260-4d8b-416f-ac07-f2433b4d5142/download/global_pcodes_adm_1_2.csv'

# Select the relevant information from the P-Codes file.
pcodes <- read_csv(url_pcode) |> 
  slice(-1) |>
  select(Location, `P-Code`, Name)

# Rename P-Codes information using lower case.
names(pcodes) <- tolower(names(pcodes))

#Redefine the data to include the named P-Code
data <- left_join(data, pcodes, join_by(ADM2_PCODE == `p-code`))
```
## Tidying
NA values in the data arise because each row or observation includes measurements over 1 month and 3 month timescales, whereas the measurement frequency is 10 days. Presenting data for different timeframes in different datasets eliminates NA values. 

The new variable "window" identifies measurements made at 10-day, 1-month and 3-month aggregation intervals. The new variable "type" identifies rainfall, average and anomaly measurements. 
```{r echo=TRUE}
rainfall <- data |>
  
  # Create new variables window and type for:
    # 10-day (rf), 1-month (r1) and 3-month (r3) measurements (window), and 
    # rainfall (h), long-term average (avg) and anomaly (q) measurements (type).
  pivot_longer(5:13, 
               names_to = "type", 
               values_to = "value") |>
  mutate(
    # Truncate the values in "type" to leave rf, r1 and r3 for 10-day, 1-month and 3-month
    # measurements.
    window = str_trunc(type, 2, ellipsis = ""),
    # Remove characters to leave h, avg, and  q for rainfall, average and anomaly measurements. 
    type = str_remove(type, "r.h_|rf|r1|r3"),
         .before = type) |>
  pivot_wider(names_from = "type") |>
  
  # Nest dataframes for cleanliness. 
  nest_by(name, n_pixels, window)
```
After data is nested by location (name), resolution (n_pixels) and measurement timeframe (window), the nested data can be cleansed for relevant measurements. No filtering is needed on the data for the 10-day (rf) timeframe because this timeframe exactly matches the frequency of measurements.
```{r echo=TRUE}
# Mapping functions to tidy nested data:
rainfall$data <-  rainfall$data |>
  
  # Filter monthly datasets for monthly rainfall measurements,
  map_at(
    grep("r1", rainfall$window), 
    filter, day(date) == 21) |> 
  
  # Filter quarterly datasets for quarterly rainfall measurements,
  map_at(
    grep("r3", rainfall$window), 
    filter, (day(date) * month(date)) %in% c(63, 126, 189, 252) == TRUE) |>
  
  # Ungroup for cleanliness.
  map(ungroup)
```
The wettest month of the year can be identified from the data. This is used as the starting month of the water-year. 
``` {r echo=TRUE}
# Tabulate monthly rainfall.
monthly <- rainfall |> group_by(name) |> filter(n_pixels == max(n_pixels) & window == "r1") |> ungroup() |> unnest(cols = c(data))

# Record the start month of the water year.
water_year_start <- month(monthly$date[which(monthly$avg == max(unique(monthly$avg)))][1])

# Plot the average and actual rainfall in all districts by month.
# Check the output saved to water_year_start for sense against the plot.
monthly |> group_by(month = factor(month(date))) |> summarise(rainfall = sum(h), average = sum(avg)) |> pivot_longer(c("rainfall", "average"), names_to = "measure", values_to = "h") |> ggplot(aes(x = month, y = h, fill = measure)) + geom_col(position = "dodge")
```

The first month of the water year is `r month.name[water_year_start]`. The chart above shows monthly rainfall (blue) and long-term average rainfall (red). 

We add a water year variable to align the beginning of the annual rainfall measurements in the data with the start of the wet season. This allows the model to capture the greatest differences between water supply and demand. 
``` {r echo=TRUE}
# Add a water year variable to align the data with water_year_start. 
rainfall$data <- 
  rainfall$data |> 
  map(mutate,
    water_year = year(date) - !(month(date) %in% water_year_start:12),
    .after = date) |>
  
  # Get cumulative rainfall for each water year in all datasets throughout each water year,
  map(group_by, water_year) |>
  map(mutate, 
      cum_h = cumsum(h), 
      .after = h) |> 
  map(ungroup, water_year)
```
## Model
To apply the model, we consider 1-month aggregated rainfall in Tororo District, Uganda at the maximum available pixel resolution.
``` {r echo=TRUE}
# Extract the data for Tororo
tororo <- rainfall |> group_by(name) |> 
  filter(
    name == "Tororo" & 
    n_pixels == max(n_pixels) & 
    window == "r1") |>
  ungroup(name) |> 
  unnest(cols = c(data))
```

Calculate monthly water balance, overflow and deficit.
``` {r echo=TRUE}
# Assume a volume of demand in cubic meters, here 10 liters per day for 1000 people (m3/year).
demand_vol <- 3 * 1000 * 364.25/1000

# Set efficiency for the capturing surface (proportion of rainfall captured).
efficiency <- 0.8

# Set area of the capturing surface (m2).
area <- 1200

# Set storage capacity (m3). Note, this should reflect the effective storage of the tank (e.g. taking into account taps and overflow pipes).
storage <- 20

# Set date of rainwater system first operation. This is only required if evaluating historic performance. To calculate probabilities for outcomes, the longest dataset available should be used.
first_operation_date <- as.Date('1981-01-21')

# Filter tororo dataframe from date of first operation.
tororo_filtered <- tororo %>%
  filter(date > first_operation_date)

# Calculate end of month water balance.
monthly_wb <- tororo_filtered %>%
  mutate(water_year = water_year,
         month = fct_inorder(month.abb[month(date)]),
         inflow = h * 0.001 * area * efficiency,
         outflow = demand_vol/12,
         balance = 0,  # Initialize balance column with 0
         unrestrained_balance = lag(balance, default = 0) + inflow - outflow,
         balance = pmax(pmin(unrestrained_balance, storage), 0),
         overflow = pmax(unrestrained_balance - storage, 0),
         deficit = pmax(-unrestrained_balance, 0))
         #.keep = "none",
         #.after = month)

```

Calculate the probability of monthly water balance scenarios (deficit, balance and overflow).
``` {r echo=TRUE}
# Calculate probability of monthly balance, overflow and deficit.
probability_of_outcomes <- monthly_wb %>%
  group_by(month) %>%
  summarise(
    f_of_balance = sum(unrestrained_balance >= 0 & unrestrained_balance <= storage),
    p_of_balance = f_of_balance / n(),
    f_of_overflow = sum(unrestrained_balance > storage),
    p_of_overflow = f_of_overflow / n(),
    f_of_deficit = sum(unrestrained_balance < 0),
    p_of_deficit = f_of_deficit / n()
  )

```

Data visualisation.
``` {r echo=TRUE}
# monthly_wb range of values by month
monthly_wb %>%
    ggplot(aes(month, unrestrained_balance)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "green") +
    geom_hline(yintercept = storage, linetype = "dashed", color = "red") +
    geom_text(aes(label = round(unrestrained_balance, 2)), vjust = -0.5, hjust = 0.5, color = "black") +  # Add data labels
    geom_smooth(method = "lm", se = FALSE) +  # Add a linear trendline
    #scale_y_continuous(sec.axis = sec_axis(~.*2, name = "h")) +
    labs(title = "Unrestrained Balance Over Months",
         x = "Month",
         y = "Unrestrained Balance",
         color = "water_year") +
    theme_bw()

# monthly_wb range of value by month as box plot
monthly_wb %>%
  ggplot(aes(x = month, y = unrestrained_balance)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "green") +
  geom_hline(yintercept = storage, linetype = "dashed", color = "red") +
  #scale_y_continuous(sec.axis = sec_axis(~.*2, name = "h")) +
  labs(title = "Distribution of Unrestrained Balance Over Months",
       x = "Month",
       y = "Unrestrained Balance",
       color = "water_year") +
  theme_bw()

# monthly_wb range of value by month, each year
monthly_wb %>%  
  ggplot(aes(month, unrestrained_balance)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "green") +
    geom_hline(yintercept = storage, linetype = "dashed", color = "red") +
    #scale_y_continuous(sec.axis = sec_axis(~.*2, name = "h")) +
    facet_wrap(~water_year) +
    scale_y_continuous(sec.axis = sec_axis(~.*2, name = "h")) +
    labs(title = "Unrestrained Balance Over Months",
         x = "Month",
         y = "Unrestrained Balance") +
    theme_bw()

# Probability of monthly water balance scenarios
    
# Reshape the data to long format
probability_of_outcomes_long <- probability_of_outcomes %>%
  pivot_longer(cols = starts_with("p_of_"), names_to = "Scenario", values_to = "Probability") %>%
  mutate(Scenario = factor(Scenario, levels = c("p_of_deficit", "p_of_balance", "p_of_overflow")))

# Create a stacked bar graph
ggplot(probability_of_outcomes_long, aes(x = month, y = Probability, fill = Scenario)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  geom_text(aes(label = paste0(round(Probability * 100), "%")), 
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white") +
  labs(x = "Month", y = "Probability (%)", fill = "Scenario") +
  ggtitle("Probabilities of Monthly Water Balance Scenarios") +
  theme_minimal()

```
## Further analysis
Further work on this model could involve:

* Refining demand projections to reflect water consumption in real settings. A linear model of demand is not realistic in a school. 
* Testing the validity of the capture surface efficiency assumption.
* Evaluating the sensitivity of the model to demand and efficiency inputs.
* Finding some way to acknowledge or predict costs. 
* Producing charts and graphs that are more appealing, intuitive or useful.
* Applying the model to other locations, perhaps using data from the sources below.

``` {r echo=TRUE}
# Rainfall data for other dry countries is available from the following locations.
url_uganda <- 
  'https://data.humdata.org/dataset/5ce9dfd7-b391-48cd-ad22-a6d0769fafe7/resource/96d0169f-0b56-428b-b59e-27e82c3f4eff/download/uga-rainfall-adm2-full.csv'
url_namibia <- 
  'https://data.humdata.org/dataset/d5211e9a-9b5b-45b5-a62c-bae4da4def46/resource/580104b3-2d95-4371-933d-9a4d1318e3c2/download/nam-rainfall-adm2-full.csv'
url_peru <- 
  'https://data.humdata.org/dataset/f5442ac6-4a5a-45a5-9763-ac60e5203bb1/resource/48fb6ece-8d03-4fde-b5e5-f5150ba207ce/download/per-rainfall-adm2-full.csv'
url_sudan <- 
  'https://data.humdata.org/dataset/139b7e9c-3c40-49e0-a44e-0eed6dad46d2/resource/eb6fbac0-1381-44a1-80e2-e49a70585f82/download/sdn-rainfall-adm2-full.csv'
url_algeria <- 
  'https://data.humdata.org/dataset/f18fb0de-5323-4c2c-a987-246ece96e8bb/resource/ecf1aa67-5c29-430c-93fc-0252ac1a2d1a/download/dza-rainfall-adm2-full.csv'
url_ethiopia <- 
  'https://data.humdata.org/dataset/423143be-315f-48d7-9e90-ae23738da564/resource/e8bb806b-c4a4-4892-aa0c-68bb7e490004/download/eth-rainfall-adm2-full.csv'
url_morocco <- 
  'https://data.humdata.org/dataset/1bd3a3a1-9c39-412b-97db-bfb80866d33f/resource/d386a362-4596-405d-8c32-ff6fa85c3632/download/mar-rainfall-adm2-full.csv'
url_egypt <- 
  'https://data.humdata.org/dataset/459b918b-1ca8-4f0a-b8b0-1e9e58829b53/resource/8e175af6-fb91-4ca0-9d25-713e262176a3/download/egy-rainfall-adm2-full.csv'
url_iran <- 
  'https://data.humdata.org/dataset/df1ba23f-d7f7-452a-bc1c-b283b8f9a958/resource/491ca0c4-a926-4690-8fe0-7b7d69cfa1c4/download/irn-rainfall-adm2-full.csv'
url_chile <- 
  'https://data.humdata.org/dataset/cb894926-763d-4912-8498-5d79a3510f60/resource/17fb3495-6340-4ce6-8c48-6e85fbb456c6/download/chl-rainfall-adm2-full.csv'
url_afghanistan <- 
  'https://data.humdata.org/dataset/3b5e8a5c-e4e0-4c58-9c58-d87e33520e08/resource/a8c98023-e078-4684-ade3-9fdfd66a1361/download/afg-rainfall-adm2-full.csv'
